Derogations, Democratic Backsliding, and International Human Rights During the COVID-19 Pandemic

### 2024-11-25 manuscript/manuscript.qmd
Address R&R issue 5

From these literatures, we draw several hypotheses:

> H~1~: States experiencing democratic backsliding will be more likely to issue derogations.
>
> H~2~: States experiencing democratic backsliding will be more likely to abuse human rights.
>
> H~3~: States experiencing democratic backsliding that also issue derogations will be less likely to abuse human rights than states that only backslide.

To test these expectations, we conduct a series of quantitative analyses followed by a qualitative examination of illustrative cases mapping on to each quadrant of @tbl-expectations. From these series of tests, we provide a deeper understanding of motivations and behavior of states during times of a health crisis and global threats to democracy.

@@

We find strong support for our hypothesis that states with a high risk of backsliding are more likely to abuse human rights. Earlier, we found that backsliding states had a higher baseline than non-backsliding states for imposing pandemic-related emergency measures. We find similar trends with human rights abuses. In all the human rights outcomes we observed, states with a high risk of backsliding have a substantially high probability of human rights abuses. For instance, states with a low risk of backsliding have more than a 90% probability of imposing discriminatory policies, while states with a high backsliding risk only have a `r preds_hr_discrim_l$high_backsliding_risk.no_derogation.none$prob_ci_nice` chance of seeing no discriminatory policies, and have high probabilities of minor, moderate, or major violations. Violations of non-derogable rights are exceptionally rare in non-backsliding states (with a 95--100% chance of no violations), while such abuses are predicted to occur a quarter to half of the time in backsliding states. States with a low risk of backsliding also see lower probabilities of abusive enforcement and media restrictions than their high-risk counterparts. Backsliding states are more likely to not specify time limits for emergency measures, though the difference is less visually striking than the other human rights outcomes in @fig-hr-preds---states at risk of backsliding are `r diffs_hr_nolimit_l$high_low_backsliding_risk.no_derogation.moderate$pp_ci_nice` percentage points (for states with no derogations) or `r diffs_hr_nolimit_l$high_low_backsliding_risk.derogation.moderate$pp_ci_nice` percentage points (for states with derogations) percentage points more likely to have no time limits than states with low risk of backsliding (`r diffs_hr_nolimit_l$high_low_backsliding_risk.derogation.moderate$p_d`). Perhaps not surprisingly, states that experienced backsliding were significantly more likely to use discriminatory policy and media restrictions during the pandemic.

Derogation status interacts with backsliding risk---among states with a high risk of backsliding, derogation is associated with better outcomes for several of the human rights violations we modeled, providing partial support for our third hypothesis. Among states at risk of backsliding, those that do not derogate have a range of probable levels of implementing discriminatory policies (@fig-hr-preds, panel A) and only a `r preds_hr_discrim_l$high_backsliding_risk.no_derogation.none$prob_ci_nice` probability of no violations, while those that do derogate have a nearly 100% chance of avoiding discriminatory policies. The difference is similar for abusive enforcement (@fig-hr-preds, panel C). States at risk of backsliding are most likely to engage in minor (≈25%) or moderate (≈33%) abusive enforcement regardless of derogation status, but the distribution of possible outcomes shifts substantially at the extremes when states derogate. The risk of major abusive enforcement drops from `r preds_hr_abusive_l$high_backsliding_risk.no_derogation.major$prob_ci_nice` to `r preds_hr_abusive_l$high_backsliding_risk.nderogation.major$prob_ci_nice` and the probability of no violations jumps from `r preds_hr_abusive_l$high_backsliding_risk.no_derogation.none$prob_ci_nice` to `r preds_hr_abusive_l$high_backsliding_risk.derogation.none$prob_ci_nice` for derogating states. The difference appears similar---though not significant---for the violation of non-derogable rights. Derogating backsliding states are roughly five percentage points less likely to violate these rights, but the difference is not significant (`r diffs_hr_ndrights_l$high_backsliding_risk.derogation_no_derogation.none$p_d`).

>>>>

From these literatures, we draw two hypotheses:

> H~1~: States experiencing democratic backsliding will be more likely to issue derogations.

> H~2~: States experiencing democratic backsliding that also issue derogations will be less likely to abuse human rights than states that only backslide.

To test these expectations, we conduct a series of quantitative analyses followed by a brief examination of illustrative cases mapping on to each quadrant of @tbl-expectations (we include more detailed case studies in the appendix). From these series of tests, we provide a deeper understanding of motivations and behavior of states during times of a health crisis and global threats to democracy.

@@

We find that states with a high risk of backsliding are more likely to abuse human rights. Earlier, we found that backsliding states had a higher baseline than non-backsliding states for imposing pandemic-related emergency measures. We find similar trends with human rights abuses. In all the human rights outcomes we observed, states with a high risk of backsliding have a substantially high probability of human rights abuses. For instance, states with a low risk of backsliding have more than a 90% probability of imposing discriminatory policies, while states with a high backsliding risk only have a `r preds_hr_discrim_l$high_backsliding_risk.no_derogation.none$prob_ci_nice` chance of seeing no discriminatory policies, and have high probabilities of minor, moderate, or major violations. Violations of non-derogable rights are exceptionally rare in non-backsliding states (with a 95--100% chance of no violations), while such abuses are predicted to occur a quarter to half of the time in backsliding states. States with a low risk of backsliding also see lower probabilities of abusive enforcement and media restrictions than their high-risk counterparts. Backsliding states are more likely to not specify time limits for emergency measures, though the difference is less visually striking than the other human rights outcomes in @fig-hr-preds---states at risk of backsliding are `r diffs_hr_nolimit_l$high_low_backsliding_risk.no_derogation.moderate$pp_ci_nice` percentage points (for states with no derogations) or `r diffs_hr_nolimit_l$high_low_backsliding_risk.derogation.moderate$pp_ci_nice` percentage points (for states with derogations) percentage points more likely to have no time limits than states with low risk of backsliding (`r diffs_hr_nolimit_l$high_low_backsliding_risk.derogation.moderate$p_d`). Perhaps not surprisingly, states that experienced backsliding were significantly more likely to use discriminatory policy and media restrictions during the pandemic.

Derogation status interacts with backsliding risk---among states with a high risk of backsliding, derogation is associated with better outcomes for several of the human rights violations we modeled, providing partial support for our second hypothesis. Among states at risk of backsliding, those that do not derogate have a range of probable levels of implementing discriminatory policies (@fig-hr-preds, panel A) and only a `r preds_hr_discrim_l$high_backsliding_risk.no_derogation.none$prob_ci_nice` probability of no violations, while those that do derogate have a nearly 100% chance of avoiding discriminatory policies. The difference is similar for abusive enforcement (@fig-hr-preds, panel C). States at risk of backsliding are most likely to engage in minor (≈25%) or moderate (≈33%) abusive enforcement regardless of derogation status, but the distribution of possible outcomes shifts substantially at the extremes when states derogate. The risk of major abusive enforcement drops from `r preds_hr_abusive_l$high_backsliding_risk.no_derogation.major$prob_ci_nice` to `r preds_hr_abusive_l$high_backsliding_risk.nderogation.major$prob_ci_nice` and the probability of no violations jumps from `r preds_hr_abusive_l$high_backsliding_risk.no_derogation.none$prob_ci_nice` to `r preds_hr_abusive_l$high_backsliding_risk.derogation.none$prob_ci_nice` for derogating states. The difference appears similar---though not significant---for the violation of non-derogable rights. Derogating backsliding states are roughly five percentage points less likely to violate these rights, but the difference is not significant (`r diffs_hr_ndrights_l$high_backsliding_risk.derogation_no_derogation.none$p_d`).
